Nail: A practical interface generator for data formats

### 2014-04-02 paper/impl.tex
github link. misc
Implementation

\paragraph{Parsing}
A generated Nail parser makes two passes through the input , the first to validate and recognise the
input, the second to bind this data to the internal model. Currently the parser is a straightforward
top down parser, although facilities have been made to add Packrat parsing to achieve linear parsing times.
\paragraph{Memory allocation} 
Many security vulnerabilities can occur when heap allocations are done improperly. Therefore, just
like Hammer, Nail avoids using the heap as much as possible, using a custom arena allocator and
allocating only fixed size blocks from the system malloc. However, Nail extends upon Hammers
approach and uses two arenas for each parsed input. One arena is used for intermediate results and is
freed(and zeroed) after parsing completes, whereas the other arena is used only for allocating the
result and has to be freed by the programmer.

\paragraph{Intermediate representation}
Most parser generators, such as Bison do not have to dynamically allocate temporary
data, as they evaluate a semantic action on every rule. However, as our goal is to perform as little
computation as possible without having validated the input, and we do not want to mix temporary
objects with the results of our parse, we need some intermediate structure to cache parse results.
This structure is also used as a value in the Packrat hash table.

@@

Under this premise, the actions need very limited information from the recogniser to correctly
handle the input stream. In particular, the parsers control flow only branches at the choice and
repeat combinators. For all other combinators, the action just has to assign a fixed sequence of
fundamental parsers to a fixed sequence of internal model fields. 
Thus we can represent the parsers success as a sequence of selected choices and counted repeats and
use this to reconstruct the syntax of the input. %TODO: This sentence is HORRIBLE. Re-work 
\paragraph{Dependency Fields}
During parsing, dependency fields occur before the context in which they are used. The parser stores
their value and retrieves it afterwards when encountering the combinator that uses them. When
generating output, the dependency field is first filled with a filler value, then later when the
first combinator that determines this fields value is encountered, the field is overwritten. Any
further combinators using this dependency will then validate that the dependency field is correct. 
\paragraph{Bootstrapping}
To demonstrate the feasibility of the nail parser generator, our parser generator uses a Nail parser
to recognise Nail grammars. A superset of the grammar language described in this paper is
implemented in 100 lines of Nail, which feed into about a thousand lines of C++ that implement the
actual parser generator. Bootstrapping is supported via an LALR implementation.

>>>>

The source code of our implementation, together with the examples described in
\S\ref{s:eval} are available on GitHub\footnote{\url{https://github.com/jbangert/nail}}.
\paragraph{Parsing}
A generated Nail parser makes two passes through the input , the first to
validate and recognise the input, the second to bind this data to the internal
model. Currently the parser is a straightforward top down parser, although
facilities have been made to add Packrat parsing to achieve linear parsing
times.
\paragraph{Memory allocation}
Many security vulnerabilities can occur when heap allocations are done
improperly. Therefore, just like Hammer, Nail avoids using the heap as much as
possible, using a custom arena allocator and allocating only fixed size blocks
from the system malloc. However, Nail extends upon Hammers approach and uses two
arenas for each parsed input. One arena is used for intermediate results and is
freed(and zeroed) after parsing completes, whereas the other arena is used only
for allocating the result and has to be freed by the programmer.

\paragraph{Intermediate representation}
Most parser generators, such as Bison do not have to dynamically allocate
temporary data, as they evaluate a semantic action on every rule. However, as
our goal is to perform as little computation as possible without having
validated the input, and we do not want to mix temporary objects with the
results of our parse, we use an append-only trace to store intermediate parser
results.

@@

Under this premise, the actions need very limited information from the
recogniser to correctly handle the input stream. In particular, the parsers
control flow only branches at the choice, repetition and constant
combinators.
Thus, for each of those combinators, we store the minimum amount of information
required to reconstruct the syntactic structure of the input. The trace is an
array of integers.
 Whenever the parser encounters a choice, it appends two integers to the trace. After it
successfully parses a choice, the parser writes the number of that choice and
the length of the trace when it began parsing that choice. When backtracking in
the input, the parser does not backtrack in the trace. This means that offsets
within the trace can be used for a Packrat hash-table to memoise backtrack-heavy
parsers.

When encountering a repetition combinator, the parser records the number of
times the inner parser succeeded and when encountering a constant parser of
variable size, it records how much input was consumed by the constant parser.

In a second pass, the parser then allocates the internal representation from an
arena allocator and binds the fields to values from the input, while following
the trace to determine how many array fields to parse and which choices to pick.
\paragraph{Dependency Fields}
During parsing, dependency fields occur before the context in which they are
used. The parser stores their value and retrieves it afterwards when
encountering the combinator that uses them. When generating output, the
dependency field is first filled with a filler value, then later when the first
combinator that determines this fields value is encountered, the field is
overwritten. Any further combinators using this dependency will then validate
that the dependency field is correct.
\paragraph{Bootstrapping}
To demonstrate the feasibility of the nail parser generator, our parser
generator uses a Nail parser to recognise Nail grammars. A superset of the
grammar language described in this paper is implemented in 100 lines of Nail,
which feed into about a thousand lines of C++ that implement the actual parser
generator. Bootstrapping is supported via an implementation of the Nail language
in the Lemon parser generator, a variant of Yacc.
